---
title: "Analysis of Covariance"
author: "Phil G Rugwiro"
date: "1/16/2021"
always_allow_html: TRUE
output:
  md_document:
    variant: gfm
    preserve_yaml: TRUE
---

```{r setup, include=FALSE}
source(here::here("rmd_config.R"))
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo= FALSE, message=FALSE, warning=FALSE}
# Installing and loading libraries that will be used in this analysis

packages = c("kableExtra","lubridate", "gridExtra", "tree", "tidyverse", "here", "caret")

# Load the package or install and load it
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)


```

Statistics is the true backbone of data science. The recent technological advancements in computational infrastructures and data analytics tools have revolutionized the way we carry out data analysis, however the fundamental statistical principles remain the foundation of everything data. 

Today, we discuss a popular statistics topic: Analysis of Covariance, commonly known by its acronym, ANCOVA. 

ANCOVA is a linear model that contains at least one covariate (continuous independent variable) and at least one categorical variable. Since the response variable is continuous, Ancova could be seen as a combination of a regression analysis (with respect to the continuous independent variable(s)) and analysis of variance (with respect to the categorical variable(s)). Another way to put this is that when we run ANCOVA, we are running regression analyses for each level in the categorical variable(s). 

For example, imagine we are trying to determine the relationship between the weight as a function of diet and age. In this example, our outcome, weight, is a continuous variable and the two independent variables are diet and age. Diet is a categorical variable, while age is a continuous variable. In this scenario, ANCOVA is used to determine the function:
<div> $$weight = f(diet, age)$$ </div>
The linear model seeks to establish a relationship between the response variable (or outcome) and the independent variables (or inputs). A model that can describe this relationship could potentially be used to make accurate data driven decisions and to predict outcomes. 


#### ANCOVA in Practice. 

This morning, as I was catching up with my mother, who is currently sheltering at home with the rest of her household due to her recent Covid-19 positive diagnosis, I got to learn that my 10 year old nephew who lives with her has recently picked up a new hobby. I was very concerned for him that he will become very frustrated and irritated by his inability to leave home, to go to school and to be around other children for the next couple of weeks. I was told not to worry about him since he's spending most of his time paying proper attention to his young 10 chickens. He used to bother his parents about money to buy more airtime and browse social media, but now he's bothering them for money to buy more chickenfeed. I was very moved by this young entrepreneur. He told me that he's only worried about lagging behind on computer lessons from school and I promised him that he will get personal tutoring from me. 
Now, in honor of my nephew's chicken farming empire, let's see ANCOVA in practice by using a built-in dataset in R, **ChickWeight**, an experiment on the effect of diet on the growth of young chicks. 

```{r}
data("ChickWeight")
```

The dataset contains four groups of chicks on different diets. Their weight measurements (grams) were taken at birth and every other day after until day 21.  

Dataset preview:



```{r echo=F}
head(ChickWeight) %>%
kbl() %>%
  kable_classic_2(full_width = F, html_font = "Arial") %>%
  kable_styling(latex_options = "scale_down")
```

The boxplot of the data shows that diet 3 is slightly better than diet 4, and that diet 2 is slightly better than diet 1. ANCOVA analysis helps determine if the true values of observed differences and whether or not they are statistically significant.

```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
ChickWeight %>%
  ggplot(aes(Diet, weight)) +
  geom_boxplot(fill='brown4', alpha = .666)
```


#### Running ANCOVA in R

Running ANCOVA in R is very simple and straightforward. 


```{r}
ancova_model <- lm(weight ~ Time+Diet, data = ChickWeight)
```

In the model above, we assumed no significant interactions between the independent variables. 

To view the results of the model, we call the summary function on the model: 

```{r}
summary(ancova_model)
```

#### Model interpretation:

- the low p_value of 2.2e-16 indicates that the model itself is very significant.
- the low p_values in the table indicate that each parameter in the model is statistically significant.
- the R-squared value of 0.7453 indicates that the model explains nearly 75% of the variations observed in the data, which is very good. 

What of the rest of the numbers in the table?

Earlier, we mentioned that when we run ANCOVA, we are in fact running regression analyses for each level in our categorical variables. In our example, we have 4 levels (diets) in our Diet variable. Therefore, we are running 4 regression analyses and require 4 intercepts and 4 slopes. 

Starting from the top (Intercept), the estimated value of 10.9244 indicates the intercept for the graph of chicks weight against time for Diet 1. 
On row 3, the estimate value of 16.1661 is used to determine the intercept for Diet 2. The actual intercept for the graph of chicks weight against time for Diet 2 is 16.1661 + 10.9244 = 27.0905. 
On row 4, the we get the intercept for Diet 3 as 36.4994 + 10.9244 = 47.4238. 
On row 5, the intercept for Diet 4 is 30.2335 + 10.9244 = 41.1579.

All 4 graphs for diets share a common slope of 8.7505 shown on row 2. 

$$weight_{diet1} = 10.9244 + 8.7505Time$$
$$weight_{diet2} = 27.0905 + 8.7505Time$$
$$weight_{diet3} = 47.4238 + 8.7505Time$$
$$weight_{diet4} = 41.1579 + 8.7505Time$$

```{r  echo=F, fig.align='center', fig.width=6, fig.height=4}
ChickWeight %>%
  ggplot(aes(Time, weight, col = Diet)) +
  geom_point() +
  geom_abline(slope = 8.7505, intercept = 10.9244, col = "brown1") +
  geom_abline(slope = 8.7505, intercept = 27.0905, col = "darkgreen") +
  geom_abline(slope = 8.7505, intercept = 47.4238, col = "cyan3") +
  geom_abline(slope = 8.7505, intercept = 41.1579, col = "blueviolet")
```

There we have it. My nephew will be advised to use Diet 4 for his chickenfeed. 

This is a simplified version of ANCOVA in R. In this example, we only had one continuous independent variable and one categorical variable with 4 levels. We ran what is called a One-Way ANCOVA. In the real world, we encounter datasets with multiple covariates and we run Multi-Way ANCOVA. In addition, we may experience interactions between independent variables. Example, imagine if Diet 2 works better from day 15 compared to days prior. That would be an indication that there exists an interaction between the Diet variable and the Time variable. In our exercise, we assumed that no such interactions exist. However, R provides means to conduct analyses with interactions.  

#### Predicting with the model

Our model above explains 75% of the variations. We can confidently use it to make predictions of the weight of chicks. 

Let's say for example that we learn of a chick that was not part of the experiment, i.e. her weight has not been recorded but we know she's been on Diet 3 the entire time. We have an improvised inspection and are required to provide the weight values of all chicks at 13 days old. Since we didn't measure the weights on day 13, we feel comfortable providing the weights from day 12 for all chicks. What do we do with this newly discovered chick? 
One acceptable way would be to estimate the weight average of all chicks on Diet 3 on day 12 and record that measurement for this new chick. 
This would be: 

```{r comment = NA}
mean(ChickWeight[ChickWeight$Time == 12,]$weight)
```
129 grams.

Another acceptable way, and perhaps more intelligent way, would be to use our model to estimate/predict the weight of this new chick. 

```{r comment = NA}
new_chick <- data.frame(Time = 12, Diet = '3')
predict(ancova_model, new_chick)
```

152 grams. 

We are likely to be more accurate by using the model estimate than average measurement. 

As I write this, my mother is in good spirits, she's determined to fight off the Covid. She still can't believe she caught the virus despite having meticulously taken all the precautions. She's known in her village as a mask-evangeliser. This truly shows the vicious nature of this virus and we must all do what we can to stop the spread. 
I will report later on the progress of my nephews chicken empire. 



