---
title: "AWS SAA"
author: "Phil G Rugwiro"
date: "1/20/2021"
output:
  md_document:
    variant: gfm
    preserve_yaml: yes
  html_document:
    df_print: paged
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source(here::here("rmd_config.R"))
```


#### Background

I set a few challenging goals for myself this year. One of them is to pass the AWS Data Analytics - Specialty certification. 

Learning something new is as much daunting as it is exhilarating. It requires discipline, commitment, and tenacity. 

Rather than jumping straight to the exam preparation for the specialty certification, I decided to first solidify my knowledge of the AWS Cloud Infrastructure. 

I had already achieved the Cloud Practitioner Certification sometime last year, so I was already feeling good about my knowledge of the AWS Cloud. However, I came to a realization that this knowledge was not to my satisfaction when I took a Solutions Architect Associate mock exam and didn't perform well. 
I was reminded that it is one thing to know the components of the infrastructure and their functions, but it is a whole other thing to know how to architect these components in an operationally excellent manner. 

I decided that I must have the SAA certification before attempting the DA specialty certification. 

#### Learning Journey

Ryan Kroonenburg from acloudguru was my instructor. His course is really a must take for all looking to gain practical knowledge of AWS. He covers the theory of each and every component of the infrastructure, complements it with lab practice and provides exam tips related to the said component. 

After completing the SAA course on acloudguru, I went back to Youtube to listen to Andrew Brown, who himself has a great free SAA course. I only did this because I really like his energy and his CCP course had previously helped me achieve a near perfect score when I took the certification exam. I let his video play in the background, which helped me with retention. 

Then I proceeded with my individual practice. I spent quite a bit of time (and some money) in the AWS Console, practicing everything I had learned, particularly, really nailing the design of VPCs and multi-tiered architectures.

#### Exam Practice

For exam practice, there is no other place to go except Whizlabs. Their SAA package contains 7 full fledged Practice Tests that mirror the actual exam. In addition to these practice exams, the package contains additional quizzes for different topics, i.e. quizzes on S3, EC2, IAM, Lambda, etc. In total, there are 760 questions to really practice for the exam. My objective was to take each of the 7 practice tests at least twice before I schedule my certification exam. Although each test has a 2h20 minutes time limit, I attempted to finish them in 90 minutes for the first take and in 45 minutes for the retake just to add more pressure on myself that is usually non-existent in a practice environment. The process is that I would take the practice test, (most likely fail it), revise my test report, and retake the test a day or two days later, then move onto the next practice test. 
The great feature on Whizlabs that I recommend is the labs that they have under many of the questions. When reviewing the test report, if there are labs associated to a given question, a link will be provided and one can complete the lab to gain even more practical experience on that topic in question. I highly recommend these labs.

To visualize my areas of weakness, I aggregated my practice test scores, for the first take and retake, grouped in all the main domains of the AWS Cloud Infrastructure: Design High Performing Architectures, Design Resilient Architectures, Design Cost-optimized Architectures, and Design Secure Applications and Architectures. This is a dataful story after all.  

Data Preview:

```{r echo=FALSE, warning=FALSE, message=FALSE}
aws_saa <- read.csv("/users/philg/desktop/aws_saa.csv")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)

head(aws_saa) %>%
kbl() %>%
  kable_classic_2(full_width = F, html_font = "Arial") %>%
  kable_styling(latex_options = "scale_down")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
aws_saa$Test.Num <- as.factor(aws_saa$Test.Num)
aws_saa$Retake.Num <- as.factor(aws_saa$Retake.Num)
```


I must admit that the practice tests made me really humble. From the graph below, the average score across all 7 test for the first try is 65, which is effectively a fail (70% is required to pass). I took the second try after a careful review of all answers and I still didn't manage to get the maximum score. I must however admit that time constraints that I put on myself played a role in the poor performance. 

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=5, fig.height=4}
aws_saa %>%
  group_by(Test.Num, Retake.Num) %>% 
  summarise(Avg = 100*sum(Correct)/sum(Number.of.Questions)) %>%
  ggplot(aes(Test.Num, Avg, col = Retake.Num)) +
  geom_point() +
  ggtitle("Practice Test Scores 1st & 2nd Tries") +
  xlab("Test Number") +
  ylab("Final Score") +
  geom_abline(intercept = 65, slope = 0, color = 'red') +
  geom_abline(intercept = 95, slope = 0, color = 'cyan3') +
  theme(plot.title = element_text(size = 10), 
        axis.title = element_text(size = 8), legend.position = "bottom") 

```
My performance was balanced across all domains of the AWS Cloud Infrastructure. No area stood out as significantly different from the rest. 

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4}
aws_saa %>%
  group_by(Domain, Retake.Num) %>%
  summarise(Avg = 100*sum(Correct)/sum(Number.of.Questions)) %>%
  ggplot(aes(Avg, Domain, fill = Retake.Num)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  ggtitle("Avg Score per Domain in 1st & 2nd Tries") +
  xlab("Average Score") +
  theme(plot.title = element_text(size = 10), 
        axis.title = element_text(size = 8), legend.position = "bottom") 
```


#### Exam day

I went into the exam, well prepared and very confident. I took my time and read all questions carefully. I gained even more confidence as the exam progressed because all the questions looked very familiar. It was still a great relief and deep satisfaction when I saw that **PASS** at the end of 140 minutes. 


#### Next steps

Now I feel ready to take on the Data Analytics Specialty Certification. If you have any good resources beyond AWS Documentation that might help accelerate my learning, please kindly share them with me. 
I hope to write a follow up post by the end of the year, reporting on the successful completion of this goal. 

### NEVER. STOP. LEARNING. 




